---
title: "baroth_OriginalHomeworkCode_01"
author: "Brooke Rothamer"
date: "2023-09-17"
output: html_document
---

---
title: "AN588-Week-2"
author: "Brooke Rothamer"
date: "2023-09-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
#Module 04

##Matrices
Matrices are collections of vectors
```{r MAtrices by column}
m <- matrix(data = c(1, 2, 3, 4), nrow = 2, ncol = 2) #Creating a matrix that fills column-wise
m
```
```{r MAtrices by  row}
n <- matrix(data = c(1, 2, 3, 4, 5, 6), nrow = 2, ncol = 3, byrow = TRUE) #Creating a matrix that fills row-wise
n
```
```{r MAtrices by binding vectors}
v1 <- c(1, 2, 3, 4)
v2 <- c(6, 7, 8, 9)
m1 <- rbind(v1, v2) #builds row-wise
m1

m2 <- cbind(v1, v2) #builds column-wise
m2
```

###Metadata about a matrix

```{r MAtrices metadata}
class(m1) #object class
dim(m1) #matrix dimensions
colnames(m1) #report column names (not defined for this matrix: NULL)
rownames(m1) #report row names
(str(m1)) #provides details about any data structure
```
##Arrays
Vectors and matrices are both types of arrays. They can have any number of dimensions but all the elements must be the same class.
```{r Arrays}
a <- array(data = 1:90, dim = c(5, 6, 3)) #creating a 3 dimensional array that is 5x6x3
a
```
##Subsetting / Indexing
Selecting elements from vectors, matrices, and arrays using their index positions in bracket notation
```{r Arrays indexing}
v <- 1:100  #This creates a vector containing all integers from 1 to 100 to use as an example
v

v[1:15] #select the first 15 elements

v[c(2, 4, 6, 8, 10)] #select a vector of specific elements

v[seq(from = 1, to = 15, by = 2)] #using a function/calculation to subset a vector
```
##Challenge 1

Creating a vector of the first sentence of *Ulysses* by James Joyce and removing punctuation.
```{r Challenge 1 vector}
library(stringr) #load in files in markdown file
Ulysses <- c("Stately, plump Buck Mulligan came from the stairhead, bearing a bowl of lather on which a mirror and a razor lay crossed.") #creating the string
Ulysses
Ulysses_unpunct <- gsub("[[:punct:]]","",Ulysses) #removing punctuation
Ulysses_unpunct
Ulysses_split <- str_split_1(Ulysses_unpunct, " ") #dividing the string into a vector of substrings. There is a space between the " " to indicate that it should split every time there is a apace. I used str_split_1 instead of str_split because str_split_1 returns a character vector which works in seq().
Ulysses_split
```
Extracting every third element of the vector using seq()
```{r Challenge 1 subset}
every3 <- Ulysses_split[seq(from = 3, to = 21, by = 3)]
every3
```

##Challenge 2
```{r Challenge 2 matrix}
mat <- matrix(data = 1:80, nrow = 8, ncol = 10, byrow = FALSE) #creating the matrix
mat
```
Given the matrix, mat, above, extract the 2nd, 3rd, and 6th columns and assign them to the variable x
```{r Challenge 2 extract 1}
x <- mat[c(2,3,6), ]
x
```
Given the matrix, mat, above, extract the 6th to 8th row and assign them to the variable y
```{r Challenge 2 extract 2}
y <- mat[c(6,8), ]
y
```
Given the matrix, mat, above, extract the elements from row 2, column 2 to row 6, column 9 and assign them to the variable z
```{r Challenge 2 extract 3}
z <- c(mat[2,2], mat[6,9])
z
```

##Challenge 3
Construct a 4-dimensional, 400 element array (5 x 5 x 4 x 4) named a consisting of the numbers 400 to 1 (i.e., a descending series)
```{r Challenge 3 4D array}
a <- array(data = 1:400, dim = c(5, 5, 4, 4))
a
```
Given this matrix, what would the following return?

a[1, 1, 1, 2]

a[2, 3, 2, ]

a[1:5, 1:5, 3, 3]
```{r Challenge 3 return values}
a[1,1,1,2] #a single point
a[2,3,2,] #a vector
a[1:5,1:5,3,3] #a matrix
```
##Overwriting
You can replace elements in a vector or matrix, or even entire rows or columns, by identifying the elements to be replaced and then assigning them new values

Starting with the matrix, mat, defined above, what will be the effects of operations below? 

```{r Overwriting a matrix}
mat[7, 1] <- 564 #replaces one point
mat[, 8] <- 2 #replaces a collumn
mat[2:5, 4:8] <- 1 #replaces a 4x5 area
mat[2:5, 4:8] <- c(20, 19, 18, 17) #replaces a 4x5 area with specific repeating values column-wise
mat[2:5, 4:8] <- matrix(data = c(20:1), nrow = 4, ncol = 5, byrow = TRUE) #replaces a 4x5 area with specific descending values row-wise
mat[, 8] <- c("a", "b") #replaces column 8 with characters "a" and "b" alternating; turns the numeric matrix into a character matric

mat
```
##Lists
Lists and data frames can be used together to group mixed structures and objects.
Lists are created using list()
```{r Lists}
s <- c("this", "is", "a", "vector", "of", "strings")  # here we have a vector
m <- matrix(data = 1:40, nrow = 5, ncol = 8)  # this is a matrix
b <- FALSE  #and this is a logical statement
l <- list(s, m, b)  # this puts them all together as a list
l
```
To extract elements from a list, use double brackets.
```{r List extracting}
l[[2]] #this will pull out element 2 which is the matrix
```
```{r List indexing 2}
l[[2]][2,6] #this will pull out position [2,6] from element 2 of the list
```
To reference/extract multiple elements from a list, you would use single bracket notation, which would itself return a list. This is called “list slicing”.
```{r List slicing}
l[1:2]  # This is taking just elements 1 and 2 from the list
```
Using str() provides details about the elements in a list:
```{r List details}
str(l)
```
Elements in lists can be named with names()
```{r List element names 1}
names(l) <- c("string", "matrix", "logical")
names(l)
```
Now, you can use the name of an item in the list to refer to it:
```{r List element names 2}
l$string  
l$matrix[3, 5]  
```
##Challenge 4
Make a nested list of a simplified primate taxonomy
```{r Challenge 4}
Lemuroidea <- c("Cheirogaleidae", "Lepilemuridae", "Indriidae", "Lemuridae",
    "Daubentoniidae")
Lorisoidea <- c("Lorisidae", "Galagidae")
Strepsirhini <- list(Lemuroidea, Lorisoidea)
names(Strepsirhini) <- list("Lemuroidea", "Lorisoidea")
Strepsirhini

Catarrhini <- c("Cercopithecidae", "Hylobatidae", "Hominidae")
Platyrrhini <- c("Cebidae", "Atelidae", "pitheciidiae")
Anthropoidea <- list(Catarrhini, Platyrrhini)
names(Anthropoidea) <- list("Catarrhini", "Platyrrhini")
Anthropoidea

Tarsiidae <- "Tarsiidae"
Tarsioidea <- list(Tarsiidae)
names(Tarsioidea) <- list("Tarsiidae")
Tarsioidea

Haplorhini <- list(Anthropoidea, Tarsioidea)
names(Haplorhini) <- list("Anthropoidea", "Tarsioidea")
Haplorhini

Primates <- list(Haplorhini, Strepsirhini)
names(Primates) <- list("Haplorhini", "Strepsirhini")
Primates
```

##Data Frames
The data.frame() command can be used to create dataframes from scratch.
```{r Data frame from scratch}
df <- data.frame(firstName = c("Nitin", "Silvy", "Ingrid", "Claire"), major = c("Biology (CM)",
    "Anthropology (Bio)", "Anthropology (Medical)", "Biology (EEB)"), sex = c("M",
    "F", "F", "F"), yearInProgram = c(1, 2, 2, 3))
df
```
Rather than creating one from scratch, we more commonly read tabular data into R from a pre-existing data file (i.e., a spreadsheet), which typically results in the table being represented as a dataframe. 
```{r data frame from file}
df <- read.csv("C:/Users/brook/OneDrive/Documents/2023_Fall/AN588-Week-2/random-people", sep = ",", header = TRUE, stringsAsFactors = FALSE)
# only print select columns
df[, c(1, 3, 4, 11, 12)]
```
```{r data frame details}
str(df)
```
You can use bracket notation to index data frames as with other structures.
```{r data frame indexing}
df[,3]
str(df[,3])
df$name.last
str(df$name.last)
df[3]
str(df[3])
df["name.last"]
str(df["name.last"])
```
You can add rows (additional cases) or columns (additional variables) to a data frame using rbind() and cbind().
```{r data frame adding and binding}
df <- cbind(df, id = c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 15, 16,
    17, 18, 19, 20))
df <- cbind(df, school = c("BU", "BU", "Harvard", "Harvard", "BU", "Boston College",
    "MIT", "BU", "BU", "UMASS Boston", "Harvard", "BU", "Boston College", "BU",
    "Harvard", "MIT", "Harvard", "BU", "UMASS Boston", "Harvard"))
df
```
cbind() caused **school** to be added as a *factor*
Using the $ operator will add **school** as a *character* vector.
```{r data frame adding using $ operator}
df$school <- c("BU", "BU", "Harvard", "Harvard", "BU", "Boston College", "MIT",
    "BU", "BU", "UMASS Boston", "Harvard", "BU", "Boston College", "BU", "Harvard",
    "MIT", "Harvard", "BU", "UMASS Boston", "Harvard")
```

##Filtering
Logical vectors may also be used to subset data frames.
```{r data frame filtering rows}
new_df <- df[df$school == "BU", ]
new_df #This new data frame only has the vectors for which BU in the School factor is TRUE.
```
```{r data frame filtering some columns}
new_df <- df[df$school == "BU", c("name.last", "name.first", "school")]
new_df #This new data frame only has the vectors for which BU in the School factor is TRUE and it only includes some columns
```

##Factors, Conversion and Coercion, and Special Data Values
You can convert factor to character data (and vice versa) using the as.character() or as.factor() commands. You can also convert/coerce any vector to a different class using similar constructs (e.g., as.numeric()), although not all such conversions are really meaningful. Converting factor data to numeric results in the the converted data having the value of R’s internal numeric code for the factor level, while converting character data to numeric results in the data being coerced into the special data value of NA (see below) for missing data.

Finally, R has three special data values that it uses in a variety of situations.

NA (for not available) is used for missing data. Many statistical functions offer the possibility to include as an argument na.rm=TRUE (“remove NAs”) so that NAs are excluded from a calculation.
Inf (and -Inf) is used when the result of a numerical calculation is too extreme for R to express
NaN (for not a number) is used when R cannot express the results of a calculation , e.g., when you try to take the square root of a negative number.

##Challenge 5
###Directions
Store the following numbers as a 5 x 3 matrix, called “m3”: 3, 0, 1 ,23, 1, 2, 33, 1, 1, 42, 0, 1, 41, 0, 2. Be sure to fill the matrix ROW-WISE.

Then, do the following:

Coerce the matrix to a data frame.

As a data frame, coerce the second column to be logical-valued

As a data frame, coerce the third column to be factor-valued

When you are done, use the str() command to show the data type for each variable in your dataframe.

###Step 1: Store the matrix
Use the matrix() function to make a matrix. Use concatonate, the c() function, to group the list of numbers as the data set that will fill the matrix. Specify the number of rows and columns. Specify that byrow=TRUE because the default is to fill by column.
```{r Challenge 5 step 1}
m3 <- matrix(data = c(3, 0, 1 ,23, 1, 2, 33, 1, 1, 42, 0, 1, 41, 0, 2), nrow = 5, ncol = 3, byrow = TRUE) #saves the matrix by the name m3
m3 #calls up the new matrix to check that it is correct
```

###Step 2: Coerce the matrix to a data frame
Use as.data.frame() to turn the matrix into a data frame and save it as a new object.
```{r Challenge 5 step 2}
df3 <- as.data.frame(m3) #saves the data frame as an object called df3
df3 #calls up the newly made data frame to check that it is correct
```

###Step 3: Coerce the second column to be logical valued
Logical valued means the data are logical arguments like TRUE and FALSE. Since column 2 is all 0s and 1s, it can be made to represent TRUE and FALSE using as.logical(). Use bracket notation [,2] to index column 2 specifically. Make sure to save the change into column 2 of the data frame, df3[,2].
```{r Challenge 5 step 3}
df3[,2] <- as.logical(df3[,2])
```

###Step 4: Coerce the third column to be factor valued
Factor valued means that each different entry in the data frame is given a level. Since column 3 contains numbers, those numeric values become factors when you use as.factor(). There are only two different values (1 and 2) so there ends up being two levels (named 1 and 2). Use bracket notation [,3] to index column 3 specifically. Make sure to save the change into column 3 of the data frame, df3[,3].
```{r Challenge 5 step 4}
df3[,3] <- as.factor(df3[,3])
```

###Step 5: Check the data type for each variable in the data frame
Use str() to report details about a structure. In this case, the structure is the data frame. Do this to check that the variables in the data frame are the object type that they are supposed to be.
```{r Challenge 5 step 5}
str(df3) #gives details about dataframe to check data type for each variable
```