---
title: "baroth_OriginalHomeworkCode_01"
author: "Brooke Rothamer"
date: "2023-09-17"
output: html_document
---

---
title: "AN588-Week-2"
author: "Brooke Rothamer"
date: "2023-09-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
#Module 04

##Matrices
Matrices are collections of vectors
```{r MAtrices by column}
m <- matrix(data = c(1, 2, 3, 4), nrow = 2, ncol = 2) #Creating a matrix that fills column-wise
m
```
```{r MAtrices by  row}
n <- matrix(data = c(1, 2, 3, 4, 5, 6), nrow = 2, ncol = 3, byrow = TRUE) #Creating a matrix that fills row-wise
n
```
```{r MAtrices by binding vectors}
v1 <- c(1, 2, 3, 4)
v2 <- c(6, 7, 8, 9)
m1 <- rbind(v1, v2) #builds row-wise
m1

m2 <- cbind(v1, v2) #builds column-wise
m2
```

###Metadata about a matrix

```{r MAtrices metadata}
class(m1) #object class
dim(m1) #matrix dimensions
colnames(m1) #report column names (not defined for this matrix: NULL)
rownames(m1) #report row names
(str(m1)) #provides details about any data structure
```
##Arrays
Vectors and matrices are both types of arrays. They can have any number of dimensions but all the elements must be the same class.
```{r Arrays}
a <- array(data = 1:90, dim = c(5, 6, 3)) #creating a 3 dimensional array that is 5x6x3
a
```
##Subsetting / Indexing
Selecting elements from vectors, matrices, and arrays using their index positions in bracket notation
```{r Arrays indexing}
v <- 1:100  #This creates a vector containing all integers from 1 to 100 to use as an example
v

v[1:15] #select the first 15 elements

v[c(2, 4, 6, 8, 10)] #select a vector of specific elements

v[seq(from = 1, to = 15, by = 2)] #using a function/calculation to subset a vector
```
##Challenge 1

Creating a vector of the first sentence of *Ulysses* by James Joyce and removing punctuation.
```{r Challenge 1 vector}
library(stringr) #load in files in markdown file
Ulysses <- c("Stately, plump Buck Mulligan came from the stairhead, bearing a bowl of lather on which a mirror and a razor lay crossed.") #creating the string
Ulysses
Ulysses_unpunct <- gsub("[[:punct:]]","",Ulysses) #removing punctuation
Ulysses_unpunct
Ulysses_split <- str_split_1(Ulysses_unpunct, " ") #dividing the string into a vector of substrings. There is a space between the " " to indicate that it should split every time there is a apace. I used str_split_1 instead of str_split because str_split_1 returns a character vector which works in seq().
Ulysses_split
```
Extracting every third element of the vector using seq()
```{r Challenge 1 subset}
every3 <- Ulysses_split[seq(from = 3, to = 21, by = 3)]
every3
```

##Challenge 2
```{r Challenge 2 matrix}
mat <- matrix(data = 1:80, nrow = 8, ncol = 10, byrow = FALSE) #creating the matrix
mat
```
Given the matrix, mat, above, extract the 2nd, 3rd, and 6th columns and assign them to the variable x
```{r Challenge 2 extract 1}
x <- mat[c(2,3,6), ]
x
```
Given the matrix, mat, above, extract the 6th to 8th row and assign them to the variable y
```{r Challenge 2 extract 2}
y <- mat[c(6,8), ]
y
```
Given the matrix, mat, above, extract the elements from row 2, column 2 to row 6, column 9 and assign them to the variable z
```{r Challenge 2 extract 3}
z <- c(mat[2,2], mat[6,9])
z
```

##Challenge 3
Construct a 4-dimensional, 400 element array (5 x 5 x 4 x 4) named a consisting of the numbers 400 to 1 (i.e., a descending series)

Given this matrix, what would the following return?
a[1, 1, 1, 2]
a[2, 3, 2, ]
a[1:5, 1:5, 3, 3]

###Step 1: Construct the array.
Use array() to construct an array. Use 400:1 to fill in the data, making sure that it descends from 400 to 1. Specify the dimensions, be sure to use c(). When R reports the array, it shows you matrices of dimensions one and two for all the combinations of the third and fourth dimensions.
```{r Challenge 3 step 1}
a <- array(data = 400:1, dim = c(5, 5, 4, 4)) #create the array
a #check that the array is correct
```

###Step 2: What do these commands return from the array?
From the array, you can call up one specific point of data by specifying all four dimensions. You can call up a vector by specifying three of the four dimensions. You can call up a matrix by specifying two of the four dimensions. You can also limit the vector or the matrix to just one section using the : function to give a range of values.
```{r Challenge 3 return values}
a[1,1,1,2] #a single point
a[2,3,2,] #a vector of values that vary in the fourth dimension
a[1:5,1:5,3,3] #a matrix but limited only to values 1:5 in the first and second dimensions
```
The point has a value of 300.
The vector contains the values 364 264 164 and 64. You can see that the same position varied only by the fourth dimension counts by 100.
The matrix gives contains the values
     [,1] [,2] [,3] [,4] [,5]
[1,]  150  145  140  135  130
[2,]  149  144  139  134  129
[3,]  148  143  138  133  128
[4,]  147  142  137  132  127
[5,]  146  141  136  131  126

##Challenge 4
###Directions
Make a nested list of a simplified primate taxonomy

###Step 1: Make a character object of the groups at the lowest level of list organization.
Within the strepsirhines, this means making objects of the lemuroidea and the lorisoidea. Use concatonate, c(), to combine the elements of the object. Since these are characters, make sure to put them in "quotations." The words lemuroidea and lorisoidea do not need to be in quotations because they are becoming the names of the objects.
```{r Challenge 4 step 1}
Lemuroidea <- c("Cheirogaleidae", "Lepilemuridae", "Indriidae", "Lemuridae",
    "Daubentoniidae")
Lorisoidea <- c("Lorisidae", "Galagidae")
str(Lemuroidea) #This is just to check that the object was made correctly
str(Lorisoidea) #This is just to check that the object was made correctly
```
### Step 2: Combine those characters into a list named for the next level up in the hierarchy.
At this stage use list() instead of c() because the objects you are combining are already objects. You also have to name the items in the list so that they aren't just named [[1]] and [[2]]. Make sure to name them in the order you added them to the list and use "quotations".
```{r Challenge 4 step 2}
Strepsirhini <- list(Lemuroidea, Lorisoidea) #Combine the elements into a list
names(Strepsirhini) <- list("Lemuroidea", "Lorisoidea") #Name the elements in the list in order and using quotations
Strepsirhini #call up the list to see that it is organized correctly
```
###Step 3: Repeat these steps for the haplorhines.
Note that in this simplified taxonomy, the anthropoids have an additional level of organization. Therefore, create the catarrhines by repeating steps 1 and 2; create the tarsids using steps 1 and 2; then combine the catarhines and tarids into the haplorhines by repeating step 2 again. Make sure to keep naming the list elements.
``` {r Challenge 4 step 3}
Catarrhini <- c("Cercopithecidae", "Hylobatidae", "Hominidae") #create the catarrhines character object 
Platyrrhini <- c("Cebidae", "Atelidae", "pitheciidiae") #create the platyrrhines character object
Anthropoidea <- list(Catarrhini, Platyrrhini) #use the list function to combine catarrhines and platyrrhines into a list
names(Anthropoidea) <- list("Catarrhini", "Platyrrhini") #name the objects in the list
Anthropoidea #check that the objects are organized correctly in the list

Tarsiidae <- "Tarsiidae" #create the tarsids character
Tarsioidea <- list(Tarsiidae) #make the tarsids character a list object (I'm not sure if this step is necessary)
names(Tarsioidea) <- list("Tarsiidae") #name the list object
Tarsioidea #check that the object is organized correctly in the list

Haplorhini <- list(Anthropoidea, Tarsioidea) #combines the anthropoid and tarsid lists into a haplorhine list
names(Haplorhini) <- list("Anthropoidea", "Tarsioidea") #name the objects in the list
Haplorhini #check that the list is organized correctly
```
###Step 4: Combine the lists into the final Primates list.
Repeating step 2 one last time, use list() to combine the Haplorhines and Strepsirhines, making sure to name the objects. Finally check that everything is in the correct order and the names match the groups that they are supposed to be naming.
```{r Challenge 4 step 4}
Primates <- list(Haplorhini, Strepsirhini) #combine the lists into the final list
names(Primates) <- list("Haplorhini", "Strepsirhini") #name the list objects
Primates #check the names and organization
```

##Challenge 5
###Directions
Store the following numbers as a 5 x 3 matrix, called “m3”: 3, 0, 1 ,23, 1, 2, 33, 1, 1, 42, 0, 1, 41, 0, 2. Be sure to fill the matrix ROW-WISE.

Then, do the following:

Coerce the matrix to a data frame.

As a data frame, coerce the second column to be logical-valued

As a data frame, coerce the third column to be factor-valued

When you are done, use the str() command to show the data type for each variable in your dataframe.

###Step 1: Store the matrix
Use the matrix() function to make a matrix. Use concatonate, the c() function, to group the list of numbers as the data set that will fill the matrix. Specify the number of rows and columns. Specify that byrow=TRUE because the default is to fill by column.
```{r Challenge 5 step 1}
m3 <- matrix(data = c(3, 0, 1 ,23, 1, 2, 33, 1, 1, 42, 0, 1, 41, 0, 2), nrow = 5, ncol = 3, byrow = TRUE) #saves the matrix by the name m3
m3 #calls up the new matrix to check that it is correct
```

###Step 2: Coerce the matrix to a data frame
Use as.data.frame() to turn the matrix into a data frame and save it as a new object.
```{r Challenge 5 step 2}
df3 <- as.data.frame(m3) #saves the data frame as an object called df3
df3 #calls up the newly made data frame to check that it is correct
```

###Step 3: Coerce the second column to be logical valued
Logical valued means the data are logical arguments like TRUE and FALSE. Since column 2 is all 0s and 1s, it can be made to represent TRUE and FALSE using as.logical(). Use bracket notation [,2] to index column 2 specifically. Make sure to save the change into column 2 of the data frame, df3[,2].
```{r Challenge 5 step 3}
df3[,2] <- as.logical(df3[,2])
```

###Step 4: Coerce the third column to be factor valued
Factor valued means that each different entry in the data frame is given a level. Since column 3 contains numbers, those numeric values become factors when you use as.factor(). There are only two different values (1 and 2) so there ends up being two levels (named 1 and 2). Use bracket notation [,3] to index column 3 specifically. Make sure to save the change into column 3 of the data frame, df3[,3].
```{r Challenge 5 step 4}
df3[,3] <- as.factor(df3[,3])
```

###Step 5: Check the data type for each variable in the data frame
Use str() to report details about a structure. In this case, the structure is the data frame. Do this to check that the variables in the data frame are the object type that they are supposed to be.
```{r Challenge 5 step 5}
str(df3) #gives details about dataframe to check data type for each variable
```